TacticToe: Learning to Prove with Tactics

### 2018-02-19 tactictoe_jv.tex
Related

The system
\textsf{ML4PG}~\cite{DBLP:journals/corr/abs-1212-3618,DBLP:journals/mics/HerasK14}
groups related proofs thanks to its clustering
algorithms. It allows \coq users to inspire themselves from similar proofs and
notice
duplicated proofs. Comparatively, our predictions comes from a much more
detailed description of the targeted goal.
%However, we simply create a single label for each tactic call whereas each of
%its
%arguments is treated independently in \textsf{ML4PG}.
%Our choice is motivated by the k-NN algorithm already used in
%\holyhammer for the a of theorems.
%, which can be easly adapted to the selection of tactics but would be improved
%by a
%version that can learn to predict tactics and their arguments independently.

\textsf{SEPIA}~\cite{DBLP:conf/cade/GransdenWR15} is a powerful system able to
generate
proof scripts from previous \coq proof examples.
Its strength lies in its ability to produce likely sequences
of tactics for solving domain specific goals. It operates by creating a model
for common sequences of tactics for a specific library.
This means that in order to propose the following tactic, only the previously
called tactics
are considered.
%intuition tells us that it should not be the main factor
Our algorithm, on the other hand, relies mainly on the characteristics of the
current goal
to decide
which tactics to apply next. In this way, our learning mechanism has to
rediscover why each
tactic was applied for the current subgoals. It may lack some useful bias for
common sequences
of tactics, but is more reactive to subtle changes. Indeed, it can be trained
on a large library and only tactics relevant to the current subgoal will be
selected.
Concerning the proof search, \textsf{SEPIA}'s %simple
breadth-first search is replaced by MCTS which allows for supervised learning
guidance in the exploration of the search tree.
Finally, \textsf{SEPIA} was evaluated on three chosen parts of the
\coq library demonstrating that it globally outperforms individual \coq
tactics. Here, we demonstrate the competitiveness of our system against
\eprover on the \holfour standard library.
%A recent improvement on this system~\cite{}\todo{Cezary: could not remember
%the
%reference that you gave me here i think it was CICM 2017} includes the
%possibility to
%substitute theorems appearing in tactics by ones belonging to the same
%cluster.
%In this work, we are also able to predict theorems appearing as arguments of
%tactics but we rely only on information from the current goal. Combining the
%two approaches by predicting from the goal and from the original tactic
%arguments is more difficult but is likely to produce more suitable lemmas.
%Proof patching: \cite{RingerYLG18}

A similar effort for stronger automation is performed in \isabelle. Nagashima
developed a proof strategy language PSL~\cite{NagashimaK17psl} which allows the
user to
program with regular tactics and combine them with tools such as
\sledgehammer~\cite{sledgehammer10} or Nitpick~\cite{Nitpick10}.
Selection of tactics from data collected in recorded proofs is in development
but for now it is up to the user to come up with a good meta-tactic.

Machine learning has also been used to advise the best library lemmas for new
ITP goals.
This can be done either in an interactive way, when the user completes the
proof based on the recommended lemmas, as in the \textsc{Mizar Proof
Advisor}~\cite{Urb04-MPTP0}, or attempted fully automatically, where such lemma
selection is handed over to the ATP component of a \emph{hammer}
system~\cite{hammers4qed,tgck-cpp15,holyhammer,BlanchetteGKKU16,mizAR40}.

@@

\section{Conclusion}\label{sec:concl}
We proposed a new proof assistant automation technique which combines
tactic-based proof search with machine learning prediction
Its implementation, \tactictoe, achieves an overall success rate of 66.4\%
on 7164 theorems of the \holfour standard library, surpassing \eprover
with auto-schedule. Its
effectiveness is especially visible on
theories which use inductive data structures, specialized decision procedures,
and custom built simplification sets.
Thanks to the learning abilities of \tactictoe, the generated proof scripts
usually reveal the high-level structure of the proof. %For these reasons,
We therefore believe that predicting ITP tactics based on the current goal
features is a very reasonable approach to automatically guiding proof search,
and that accurate predictions can be obtained by learning from the knowledge
available in today's large formal proof corpora.

>>>>

\textsf{ML4PG}~\cite{DBLP:journals/corr/abs-1212-3618,DBLP:journals/mics/HerasK14}
groups related proofs using clustering
algorithms. It allows \coq users to inspire themselves from similar proofs and
notice
duplicated proofs. Comparatively, our predictions comes from a much more
detailed description of the targeted goal.\todo{and can be used as an ATP}
%However, we simply create a single label for each tactic call whereas each of
%its
%arguments is treated independently in \textsf{ML4PG}.
%Our choice is motivated by the k-NN algorithm already used in
%\holyhammer for the a of theorems.
%, which can be easly adapted to the selection of tactics but would be improved
%by a
%version that can learn to predict tactics and their arguments independently.

\textsf{SEPIA}~\cite{DBLP:conf/cade/GransdenWR15} can
generate
proof scripts from previous \coq proof examples.
Its strength lies in the ability to produce likely sequences
of tactics for solving domain specific goals. It operates by creating a model
for common sequences of tactics for a specific library.
This means that in order to propose the next tactic, only the previously
called tactics
are considered.
%intuition tells us that it should not be the main factor
Our algorithm, on the other hand, relies mainly on the characteristics of the
current goal
to decide
which tactics to apply next. In this way, our learning mechanism has to
rediscover why each
tactic was applied for the current subgoals. It may lack some useful bias for
common sequences
of tactics, but is more reactive to subtle changes. Indeed, it can be trained
on a large library and only tactics relevant to the current subgoal will be
selected.
Concerning the proof search, \textsf{SEPIA}'s %simple
breadth-first search is replaced by MCTS which allows for supervised learning
guidance in the exploration of the search tree.
Finally, \textsf{SEPIA} was evaluated on three chosen parts of the
\coq library demonstrating that it globally outperforms individual \coq
tactics. Here, we demonstrate the competitiveness of our system against
\eprover on the \holfour standard library.
%A recent improvement on this system~\cite{}\todo{Cezary: could not remember
%the
%reference that you gave me here i think it was CICM 2017} includes the
%possibility to
%substitute theorems appearing in tactics by ones belonging to the same
%cluster.
%In this work, we are also able to predict theorems appearing as arguments of
%tactics but we rely only on information from the current goal. Combining the
%two approaches by predicting from the goal and from the original tactic
%arguments is more difficult but is likely to produce more suitable lemmas.
%Proof patching: \cite{RingerYLG18}

\todoi{maybe the paragraph can be generally proof strategy languages including ltac and eisbach}
A similar effort for stronger automation is performed in \isabelle. Nagashima
developed a proof strategy language PSL~\cite{NagashimaK17psl} which allows the
user to
program with regular tactics and combine them with tools such as
\sledgehammer~\cite{sledgehammer10} or Nitpick~\cite{Nitpick10}.
Selection of tactics from data collected in recorded proofs is in development
but for now it is up to the user to come up with a good meta-tactic.

Machine learning has also been used to advise the best library lemmas for new
ITP goals.
This can be done either in an interactive way, when the user completes the
proof based on the recommended lemmas, as in the Mizar proof advisor
\textsc{MizAR}~\cite{Urb04-MPTP0,mizAR40}, or attempted fully automatically, where such lemma
selection is handed over to the ATP component of a \emph{hammer}
system~\cite{hammers4qed,tgck-cpp15,holyhammer,BlanchetteGKKU16}.

@@

\section{Conclusion}\label{sec:concl}
We proposed a new proof assistant automation technique which combines
tactic-based proof search with machine learning prediction.
Its implementation, \tactictoe, achieves an overall success rate of 66.4\%
on 7164 theorems of the \holfour standard library, surpassing \eprover
with auto-schedule. Its
effectiveness is especially visible on
theories which use inductive data structures, specialized decision procedures,
and custom built simplification sets.
Thanks to the learning abilities of \tactictoe, the generated proof scripts
often reveal the high-level structure of the proof. %For these reasons,
We therefore believe that predicting ITP tactics based on the current goal
features is a very reasonable approach to automatically guiding proof search,
and that accurate predictions can be obtained by learning from the knowledge
available in today's large formal proof corpora.
